---
title: "Lab 3 (Review of Lab 2-2): Data Frames and filter()"
author: "April Truong"
date: "Winter 2019, January 25"
output: 
  html_document: 
    keep_md: yes
---
 
 
## Beginning of Class

Always check directory
```{r}
getwd()
```

Make sure to call/load tidyverse; ignore whatever is outputted
```{r}
library("tidyverse")
```


## Reading the fish data

Begin by pulling up your data
  Copy and past working directory into the read_csv()
  Separated by /
Check Environment for the data
```{r}
fish <- readr::read_csv("C:/Users/Apple/Desktop/aptruong/data/Gaeta_etal_CLC_data.csv")
```


## Summary Functions

Output column names; remind self what you're working with
```{r}
names(fish)
```

```{r}
nrow(fish) #the number of rows
```

```{r}
ncol(fish) #the number of columns
```

```{r}
dim(fish) #total dimensions
```

```{r}
colnames(fish) #column names
```

First 6 
Beneath each col names is class of data 
    dbl - numeric
    chr - character
    
```{r}
head(fish)
```

`summary()` and `str()` are classic functions used by many R programmers. `glimpse()` is part of the tidyverse.

- Summary (don't confuse with summarize)
  Give statistics to go along with each col
```{r}
summary(fish)
```

- str
```{r}
str(fish)
```

- Glimpse
```{r}
glimpse(fish)
```

## Practice: Data Frames (from Lab 2-2 Review)

1. Load the data `mammal_lifehistories_v2.csv` and place it into a new object called `mammals`.
```{r}
mammals <- readr::read_csv("C:/Users/Apple/Desktop/aptruong/data/mammal_lifehistories_v2.csv")

```

2. Provide the dimensions of the data frame.
```{r}
dim(mammals) #total dimensions
```

3. Display the column names in the data frame. 
```{r}
colnames(mammals)
```

4. Use str() to show the structure of the data frame and its individual columns; compare this to glimpse(). 
```{r}
str(mammals)
```

5. Print out the first few rows of the data using the function head().  
```{r}
head(mammals)
```


## Dplyr: filter()

    Interested in smaller subset of the data. In order to extract data from dataframe, use filter()

First thing: keep an idea of colnames
```{r}
colnames(fish)
```

What filters do you want to filter by? 
If only interested in L fish, use filter to only filter those
    Note: make sure to use ==, R distinguises b/t the two
```{r}
filter(fish, lakeid =="AL")
```
Now given new data frame restricted to AL lakes

Let's filter more, by numeric instead of character: pull rows where scale legnth are > or = to 350
```{r}
filter(fish, length >= 350 )
```
Reminder: not storing this data yet

Combine filters
```{r}
filter(fish, length == 350 & lakeid == "AL")
```

Using AND and OR
```{r}
filter(fish, length == 167 & length == 175)
```
Think: are there fish with 167 length? 175 length? Well yea, but this is asking for a fish that has BOTH 167 and 175 length. So use OR '|'
```{r}
filter(fish, length == 167 | length == 175)
```

Using '!'to filter for NOT
```{r}
filter(fish, !length == 175)
```


## Practice: filter() (from Lab 2-2 Review)

1. Filter the `fish` data to include the samples from lake `DY`.
```{r}
filter(fish, lakeid == "DY")
```

2. Filter the data to include all lakes except AL.
```{r}
filter(fish, !lakeid == "AL")
```

3. Filter the data to include all lakes except AL and DY.
```{r}
filter(fish, !lakeid == "AL" & !lakeid == "DY")
```

4. Filter the data to include all fish with a scale length greater than or equal to 11.
```{r}
filter(fish, scalelength >= 11)
```

5. Filter the data to include fish only from lake AL and with a scalelength greater than or equal to 2 and less than or equal to 4.
```{r}
fish%>%
  select(lakeid, scalelength)%>%
  filter(lakeid=="AL")%>%
  filter(scalelength >= 2 & scalelength <= 4)
```

Another way to do this is to separate by commas (create independent filter operations)
```{r}
filter(fish, lakeid =="AL",
       scalelength >= 2 & scalelength <= 4)
```
Note: the '&' restricts it to ONLY scalelengths of 2 and 4


## Pipes (%>%)
  Super helpful b/c makes codes cleaner
  Think about "flowing data into another command"     
    #supermariobros
  Shortcut: ctrl + shift + m

```{r}
fish %>% #take fish and pipe it to function filter()
  filter(lakeid == "AL")

```

## Select (select())
```{r}
fish %>% 
  select(lakeid, scalelength)
```
Now combine pipe & select()
```{r}
fish %>%
  select(lakeid, scalelength)%>%
  filter(lakeid=="AL")%>%
  filter(scalelength >= 2 & scalelength <= 4)
```
Helpful for big data!

## How  to store as independent dataframe
  Assign to new object
```{r}
fish_new <- fish %>%
  select(lakeid, scalelength)%>%
  filter(lakeid=="AL")%>%
  filter(scalelength >= 2 & scalelength <= 4)
```



## Recap

Should feel comfortable with 
  - creating .md file
  - filtering
  
Make sure to clean up notes, can post to github so you can have access online (esp if can't have own laptop)
  
Can now do HW#2








