---
title: "Lab 2 Homework"
author: "April Truong"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: cerulean
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code, keep track of your versions using git, and push your final work to our [GitHub repository](https://github.com/FRS417-DataScienceBiologists). I will randomly select a few examples of student work at the start of each session to use as examples so be sure that your code is working to the best of your ability.

## Load the tidyverse
```{r}
library("tidyverse")
```

## Mammals Sleep
For this assignment, we are going to use built-in data on mammal sleep patterns.  

1. From which publication are these data taken from? Don't do an internet search; show the code that you would use to find out in R.

```{r}
# read the data first
msleep

# information
?msleep
```
> Publication from National Academy of Sciences

2. Provide some summary information about the data to get you started; feel free to use the functions that you find most helpful.

```{r}
str(msleep)
```
> Note: other functions include 'summary()' and 'glimpse()'(which is part of the tidyverse)

```{r}
## other functions

#names(msleep)

#nrow(msleep) #the number of rows

#ncol(msleep) #the number of columns

#dim(msleep) #total dimensions

#colnames(msleep) #column names

#head(msleep) #first 6 

#tail(msleep) #last 6
```


3. Make a new data frame focused on body weight, but be sure to indicate the common name and genus of each mammal. Sort the data in descending order by body weight.
```{r}
# practice with pipes, wish I was playing Smash Ultimate instead

IRL_yoshi_bodywt <-
  msleep %>%
  select (name, genus, bodywt)

IRL_yoshi_bodywt
```

```{r}
# sort in descending order

IRL_yoshi_bodywt[order(-IRL_yoshi_bodywt$bodywt),]
```


4. We are interested in two groups; small and large mammals. Let's define small as less than or equal to 1kg body weight and large as greater than or equal to 200kg body weight. For our study, we are interested in body weight and sleep total.
Make two new objects (large and small) based on these parameters. Sort the data in descending order by body weight.

```{r}
# small group

small_group <- msleep %>%
  select(sleep_total, bodywt)%>%
  filter(bodywt <= 1) 
small_group

small_group[order(-small_group$bodywt),]
```

```{r}
# large group

large_group <- msleep %>%
  select(sleep_total, bodywt)%>%
  filter(bodywt >= 200) 
large_group

large_group[order(-large_group$bodywt),]
```


5. Let's try to figure out if large mammals sleep, on average, longer than small mammals. What is the average sleep duration for large mammals as we have defined them?

```{r}
mean(large_group$sleep_total)
```
> Average sleep duration for large mammals around 3 hours

6. What is the average sleep duration for small mammals as we have defined them?

```{r}
mean(small_group$sleep_total)
```
> Average sleep duration for small mammals almost 13 hours

7. Which animals sleep at least 18 hours per day? Be sure to show the name, genus, order, and sleep total. Sort by order and sleep total.

```{r}
msleep %>%
  select (name, genus, order, sleep_total)%>% 
  filter (sleep_total>=18) %>%
  arrange (order, sleep_total)
```
> five animals sleep at least 18 hours per day: big brown bat, little brown bat, giant armadillo, north american opossum, thick-tailed opposum

## Push your final code to [GitHub](https://github.com/FRS417-DataScienceBiologists)
Make sure that you push your code into the appropriate folder. Also, be sure that you have check the `keep md` file in the knit preferences.