---
title: "Lab 5 Homework"
author: "April Truong"
date: "Winter 2019"
output:
  html_document:
    keep_md: yes
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code, keep track of your versions using git, and push your final work to our [GitHub repository](https://github.com/FRS417-DataScienceBiologists). I will randomly select a few examples of student work at the start of each session to use as examples so be sure that your code is working to the best of your ability.

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Mammals Life History
Let's revisit the mammal life history data to practice our `ggplot` skills. Some of the tidy steps will be a repeat from the homework, but it is good practice. The [data](http://esapubs.org/archive/ecol/E084/093/) are from: *S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.*

1. Load the data.
```{r message=FALSE, warning=FALSE}
life_history <- readr::read_csv("C:/Users/Apple/Desktop/aptruong/data/mammal_lifehistories_v2.csv")
```

2. Use your preferred function to have a look. Do you notice any problems?
```{r}
#install.packages("skimr")
library("skimr")
life_history %>% 
  skimr::skim()
```
> yes, there are problems: NA's

3. There are NA's. How are you going to deal with them?
```{r}
life_history_replace <- 
  life_history %>% 
  na_if("-999") #replace x data with NA
life_history_replace
```


4. Where are the NA's? This is important to keep in mind as you build plots.
```{r}
msleep %>% 
  summarize(number_nas= sum(is.na(life_history_replace))) 
#total of 4977 NAs
life_history_replace %>% 
  purrr::map_df(~ sum(is.na(.))) %>%
  gather(variable, value = "number_nas") %>%
  arrange(desc(number_nas))
#most NAs from wean_mass
```


5. Some of the variable names will be problematic. Let's rename them here as a final tidy step.

```{r}
life_history_replace <- 
  life_history_replace %>% 
  dplyr::rename(
          genus        = Genus,
          wean_mass    = `wean mass`,
          max_life     = `max. life`,
          litter_size  = `litter size`,
          litters_yr   = `litters/year`
          )
```

##`ggplot()`
For the questions below, try to use the aesthetics you have learned to make visually appealing and informative plots. Make sure to include labels for the axes and titles.
```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

6. What is the relationship between newborn body mass and gestation? Make a scatterplot that shows this relationship. 
```{r}
life_history_replace %>% 
  ggplot(aes(x=newborn, y=gestation))+
  geom_jitter()+
  labs(title = "Newborn Body Mass vs. Gestation",
       x = "Newborn Mass (g)",
       y = "Gestation (months)")+ 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```


7. You should notice that because of the outliers in newborn mass, we need to make some changes. We didn't talk about this in lab, but you can use `scale_x_log10()` as a layer to correct for this issue. This will log transform the y-axis values.
```{r}
life_history_replace %>% 
  ggplot(aes(x=newborn, y=gestation))+
  scale_x_log10()+
  geom_jitter()+
  labs(title = "Newborn Body Mass vs. Gestation: scaled log10",
       x = "Newborn Mass (g)",
       y = "Gestation (months)")+ 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```


# 8,9,10 optional

8. Now that you have the basic plot, color the points by taxonomic order.
```{r}
life_history_replace %>% 
  ggplot(aes(x=newborn, y=gestation, size=mass, color=order))+
  scale_x_log10()+
  geom_jitter()+
  labs(title = "Newborn Body Mass vs. Gestation: scaled log10",
       x = "Newborn Mass (g)",
       y = "Gestation (months)")+ 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```


#9. Lastly, make the size of the points proportional to body mass.
```{r}
life_history_replace %>% 
  ggplot(aes(x=newborn, y=gestation, group=order, fill=order, color=order, size=order))+
  scale_x_log10()+
  geom_jitter()+
  labs(title = "Newborn Body Mass vs. Gestation: scaled log10",
       x = "Newborn Mass (g)",
       y = "Gestation (months)")+ 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
```


#10. Make a plot that shows the range of lifespan by order.
```{r}
life_history_replace %>% 
  ggplot(aes(x=order, y=max_life, group=order, fill=order))+ 
  geom_boxplot()+
  scale_y_log10()+
  labs(title = "Range of Lifespan by Order",
       x = "Order",
       y = "Lifespan (months)")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))
```


## Push your final code to [GitHub](https://github.com/FRS417-DataScienceBiologists)
Make sure that you push your code into the appropriate folder. Also, be sure that you have check the `keep md` file in the knit preferences.